Unitary evolution as a quasistatic process of infinitesimal collapses

### 2019-09-18 Articles/2019-ForQuantumOptics/article.tex
Moved all text back to article

\begin{abstract}
Any state, to be properly defined, has to be an equilibrium of all processes happening at a faster timescale. As processes with equilibria are idempotent, they can be expressed as projections in a probability space. Assuming a quasistatic evolution, then, means the equilibria of the faster scale processes will evolve slowly, giving rise to continuous and unitary evolution. These ideas are sufficient to get the general form of the projection postulate and the time evolution of quantum mechanics. Violations of the Schroedinger equation, then, should be expected only for processes that cannot be considered quasistatic.
\end{abstract}

@@

In an effort to better understand why the frameworks of classical and quantum mechanics are what they are, we have worked on identifying a minimal set of physical assumptions from which both can be derived.[cite] This gave us a new perspective that is somewhat unusual among people who work on fundamental questions in physics. The typical view is that the theories are so different that they require a radically different set of starting points, and that classical mechanics in particular should only be viewed as an approximation of quantum mechanics. In this view, the similarities between the theories stem from one being an approximation of the other.

@@

In this paper we outline the fundamental elements of a generic model for a physical system, their relationships and, in particular, how they work in the context of quantum mechanics. The general idea is that for a state to be experimentally prepared, manipulated and measured, it needs to be ``stable'' for what may be a very short but certainly finite amount of time. That is, it must be a, possibly metastable, equilibrium of faster scale processes. But a process for which all outputs are equilibria is idempotent (i.e. gives the same output no matter how many times it is applied) and is therefore mathematically modeled by a projection operation on the space of statistical distributions. We identify quantum projections as these processes. If we want a deterministic law, then, it must be a quasistatic evolution in terms of those faster scale processes. As the change must be small, this leads to continuous evolution over the state space and unitary evolution in the space of statistical distributions. We identify unitary evolution as this deterministic law.

@@

@@

If we accept this premise, then the natural thing to do is to write the conditions for a quasistatic deterministic evolution. That is, we assume that at each moment the system is in equilibrium and we change the state just a little so that each pure state is mapped to a single well-defined pure state. That is, if $\psi_{t}$ and $\psi_{t+dt}$ are two states infinitesimally close in time and $\mathcal{p}_{t+dt}$ is the process that between $[t, t+dt]$ transitions one into the other, then every time $\psi_{t}$ is the input of that process $\psi_{t+dt}$ is its output. Mathematically:
\begin{equation}
\begin{aligned}
\mathcal{p}_{t+dt}(\iota(\psi_{t}))&=\iota(\psi_{t+dt}) \\
\end{aligned}
\end{equation}
Applying the definitions we find:
\begin{equation}
\begin{aligned}
\mathcal{p}_{t+dt}(\iota(\psi_{t}))(\psi_{t+dt}) &=
\frac{\braket{\psi_{t} | \psi_{t+dt}} \braket{\psi_{t+dt}|\psi_{t}}}{\braket{\psi_{t}|\psi_{t}}} = 1 \\
\end{aligned}
\end{equation}
This requires the evolution to be unitary and we can use standard techniques from here, which will lead to the Schroedinger equation. In other words, \textbf{the Schroedinger equation is the quasistatic deterministic evolution of a process of continuous projections}. This approach, therefore, proceeds in the opposite way of others: instead of assuming the Schroedinger equation is always valid and deriving projections as a special case, we assume projections are valid approximations of faster scale processes and the Schroedinger equation is the special quasistatic deterministic case.

We can briefly see why this approach makes intuitive sense. The closer two states are in the Hilbert space, the greater the chance that one will transition to the other during a projection. Continuous evolution, then, means small deterministic changes. Unitary evolution, in particular, conserves entropy, which is also the hallmark of deterministic and reversible evolution. On the other hand, a projection can increase entropy, which is the hallmark of a transition that went out of equilibrium and is not reversible. Since any two unitary vectors can be connected by a unitary operation (i.e.~a rotation), it makes sense that all states are equilibria since they can all be connected by a deterministic and reversible process and that they all carry the same entropy. We also note that in quantum field theory, to calculate cross-sections and particle decays, one puts the initial states at minus infinity and the final states at plus infinity for processes that, in reality, take a tiny fraction of a second. This is reminiscent of a quasistatic approximation. Finally, while the approach presented here does not solve the so called ``measurement problem'', it does however give us an understanding of why we have the problem to begin with: our states are really just the equilibrium states of faster scale processes and those cannot, in general, be described by quasistatic ones.

%The connection to measurement is simply that, as we will see later in more detail, such processes are always required if we want to measure a quantity, even in classical mechanics or thermodynamics. For example, if we want to measure the temperature of a gas it needs to be in equilibrium for long enough for us to conduct the measurement. If we are measuring the position of a ball, we need to integrate over a (small) period of time to integrate the giggling created by the random collisions of air particles. That is: if we want to measure the vertical spin component of an electron, something needs to make sure that we actually have and electron and its vertical component is well defined. So we need a process whose outputs are equilibria with those properties.

@@

Every process that has equilibria is characterized by a time scale needed to reach equilibrium. That time will set the maximum time resolution we can use with that choice of states and processes. To go to faster time scales we inevitably need to change both states and processes. For example, as we change the volume of a gas, the region against the moving wall will be compressed sooner than the opposite region, but if we wait a short while we will again obtain a uniform pressure. If we want to study that transition, we may conceptually divide the system into small volumes and the pressure is replaced by a pressure field. Still, we need to assume each small volume to have enough molecules to be in equilibrium. We can go into more detail, and describe the distribution of position and velocities of the molecules. Yet, we need to assume that the molecules themselves are not undergoing chemical changes. We could study the atoms of the molecules, yet we would have to assume that the atoms are stable under nuclear fission. We could study the electrons and nucleons of the atoms, yet we would have to assume they are stable under weak nuclear forces. As we move to a finer and finer description, we have faster and faster scale processes that we assume do not change the basic constituents. At each level of the description what is assumed to be in equilibrium changes and what the laws describe changes as well. Yet, the overall structure remains the same: we have a timescale $dt$ such that within each interval $[t-dt, t]$ there is a process $\mathcal{p}_t$ such that the state $\mathcal{s}_t$ is an equilibrium of $\mathcal{p}_t$. The evolution $\mathcal{E}$, then, is quasistatic in terms of the processes in $\mathcal{P}$. That is, \textbf{all deterministic laws are quasistatic evolutions in terms of system-environment interaction processes}.

We have reached, then, the second major point of this article: \textbf{the mathematical structure we used to rearrange quantum mechanics is a general structure that applies to all systems} and as such we call it the \emph{fundamental model of physics}. It is fundamental because it always applies and what changes case by case are the specifics. If we study a ball using classical particle mechanics, for example, the position and momentum of the center of mass will be defined in a short time scale integrating the jiggling created by the random collisions of air particles. That averaging process will be the same regardless of the values of position and momentum; all states are equilibria of the same process and therefore $\mathcal{P}$ contains only one element: the identity map. Therefore, while the overall structure still exists, it is mathematically trivial and can be effectively disregarded as we concentrate on the law of evolution exclusively.

@@

We have seen that quantum mechanics fits within a more general framework that applies to all physical theories. States are always equilibria of faster scale processes, which in quantum mechanics are modeled with the projection. States are chosen to satisfy a deterministic law of evolution, necessarily describing a quasistatic process, which in quantum mechanics is modeled by the Schroedinger equation. This general framework only uses standard physics concepts and puts no special role on the observer except the choice of what processes the system will be subjected to.

Violations of the Schroedinger equation, then, should be expected only for processes that cannot be considered quasistatic. If we were to implement a measurement process, then, in a way that the transfer of information between the system and the probe happens in a slow and controlled fashion, this could not provide a violation of the Schroedinger equation, nor would it tell us that all projections are quasistatic processes. A better description of the faster scale processes given by the projections, then, will inevitably need a new state space as the standard quantum states only describe equilibria, and we should expect the faster scale processes to go out of equilibrium. Ultimately, a description at Plank time scale will fail to be continuous and quasistatic since a finite change in a finite time will not be divided into infinitesimal adiabatic changes over infinitely many time steps.

To understand the suitable mathematical framework for those regimes where unitary evolution may not hold, we are proceeding by first identifying the physical requirements for all mathematical structures already used in physics, starting from the most basic ones like topological spaces (upon which is based all of differential geometry) and $\sigma$-algebras (upon which is based all of measure theory and probability theory). This long term effort is centered mainly around the development of an open access book in which these issues can be explored in full detail and with a rigor more typical of mathematical disciplines. This rigor forces us to spell out the often hidden assumptions we have while constructing physical theories. By making explicit, for example, under which physically motivated circumstances time can be modeled by real numbers, or when states can be represented within Hilbert spaces, we believe we will be in a better position to understand what happens when those assumptions fail.

@@

\bibliography{bibliography}

>>>>

\begin{abstract}
Any state, to be properly defined, has to be an equilibrium of all processes happening at a faster timescale. As processes with equilibria are idempotent, they can be expressed as projections in a probability space. Assuming a quasistatic evolution, then, means the equilibria of the faster scale processes will evolve slowly, giving rise to continuous and unitary evolution. In this paper we show that these ideas are sufficient to get the general form of the projection postulate and the time evolution of quantum mechanics, while also creating a model that applies to other physical theories, such as classical mechanics and thermodynamics. In this light, the different theories are specializations of a broader model. As unitary evolution corresponds to quasistatic evolution, violations of the Schr\"{o}dinger equation, then, should be expected only for processes that cannot be considered quasistatic.
\end{abstract}

@@

In an effort to better understand why the frameworks of classical and quantum mechanics are what they are, we have worked on identifying a minimal set of physical assumptions from which both can be derived.\cite{Carc1} This gave us a new perspective that is somewhat unusual among people who work on fundamental questions in physics. The typical view is that the theories are so different that they require a radically different set of starting points, and that classical mechanics in particular should only be viewed as an approximation of quantum mechanics. In this view, the similarities between the theories stem from one being an approximation of the other.

@@

In this paper we outline the fundamental elements of a general model for a physical system, their relationships and, in particular, how they work in the context of quantum mechanics. The basic idea is that for a state to be experimentally prepared, manipulated and measured, it needs to be ``stable'' for what may be a very short but certainly finite amount of time. That is, it must be a, possibly metastable, equilibrium of faster scale processes. But a process for which all outputs are equilibria is idempotent (i.e. gives the same output no matter how many times it is applied) and is therefore mathematically modeled by a projection operation on the space of statistical distributions. We identify quantum projections as these processes.Quantum projections, then, should be considered in the same spirit as out-of-equilibrium thermodynamic processes: the dynamics is taken as a black box with no further characterization, there is only a clear initial and final state, and the process can, in general, increase entropy. Quantum states and projections are taken as prime notions in the same way that equilibrium is a prime notion in thermodynamics. If we want a deterministic law, then, it must be a quasistatic evolution in terms of those faster scale processes. As the change must be small, this leads to continuous evolution over the state space and unitary evolution in the space of statistical distributions. We identify unitary evolution as this deterministic law.

@@

It should be noted that while the objective is to put together a general mathematical model that can fit not only quantum mechanics, but other physical theories as well, the model itself cannot be proven mathematically, in the same way that the axioms of set theory, or any other formal system, cannot be proven. Moreover the mathematical structure we want to impose is necessarily ``simple" because simpler structures are the more general ones (i.e. every mathematical object is an element of a set, the simplest structure). Therefore the bulk of the work is, appropriately, to present a physical justification as to why the simple mathematical structure we lay out is necessary.

@@

If we accept this premise, then the natural thing to do is to write the conditions for a quasistatic deterministic evolution. That is, we assume that at each moment the system is in equilibrium and we change the state just a little so that each pure state is mapped to a single well-defined pure state. That is, if $\psi_{t}$ and $\psi_{t+dt}$ are two states infinitesimally close in time and $\mathcal{p}_{t+dt}$ is the process that between $[t, t+dt]$ transitions one into the other, then every time $\psi_{t}$ is the input of that process $\psi_{t+dt}$ is its output. Mathematically:
\begin{equation}
\begin{aligned}
\mathcal{p}_{t+dt}(\iota(\psi_{t}))&=\iota(\psi_{t+dt}) \\
\end{aligned}
\end{equation}
Applying the definitions we find:
\begin{equation}
\begin{aligned}
\mathcal{p}_{t+dt}(\iota(\psi_{t}))(\psi_{t+dt}) &=
\frac{\braket{\psi_{t} | \psi_{t+dt}} \braket{\psi_{t+dt}|\psi_{t}}}{\braket{\psi_{t}|\psi_{t}}} = 1 \\
\end{aligned}
\end{equation}
This requires the evolution to be unitary and we can use standard techniques from here, which will lead to the Schr\"{o}dinger equation. In other words, \textbf{the Schr\"{o}dinger equation is the quasistatic deterministic evolution of a process of continuous projections}. This approach, therefore, proceeds in the opposite way of others: instead of assuming the Schr\"{o}dinger equation is always valid and deriving projections as a special case, we assume projections are valid approximations of faster scale processes and the Schr\"{o}dinger equation is the special quasistatic deterministic case.

We can briefly see why this approach makes intuitive sense. The closer two states are in the Hilbert space, the greater the chance that one will transition to the other during a projection. Continuous evolution, then, means small deterministic changes. Unitary evolution, in particular, conserves entropy, which is also the hallmark of deterministic and reversible evolution. On the other hand, a projection can increase entropy, which is the hallmark of a transition that went out of equilibrium and is not reversible. Since any two unitary vectors can be connected by a unitary operation (i.e.~a rotation), it makes sense that all states are equilibria since they can all be connected by a deterministic and reversible process and that they all carry the same entropy. We also note that in quantum field theory, to calculate cross-sections and particle decays, one puts the initial states at minus infinity and the final states at plus infinity for processes that, in reality, take a tiny fraction of a second. This is reminiscent of a quasistatic approximation.

Finally, we note that the approach presented here does not solve the so called ``measurement problem''\cite{Genovese,Bassi}, at least not in the way it is typically framed. We still do not have a description of what happens during an arbitrary process (e.g. a projection), yet we have an explanation of why we cannot provide one in terms of a quasistatic evolution (i.e. unitary evolution) since a special case cannot describe the general one.\footnote{We purposely leave unspecified whether the faster scale processes are driven by internal or external dynamics, or by a mixture of the two. As such, this approach does not technically pick an interpretation: it merely states that whatever happens during a measurement in a given interpretation is also happening at every instant in time.}

@@

Every process that has equilibria is characterized by a time scale needed to reach equilibrium. That time will set the maximum time resolution we can use with that choice of states and processes. To go to faster time scales we inevitably need to change both states and processes. For example, as we change the volume of a gas, the region against the moving wall will be compressed sooner than the opposite region, but if we wait a short while we will again obtain a uniform pressure. If we want to study that transition, we may conceptually divide the system into small volumes and we replace the pressure by a pressure field. Still, we need to assume each small volume has enough molecules to be in equilibrium. We can go into more detail, and describe the distribution of position and velocity of the molecules. Yet, we need to assume that the molecules themselves are not undergoing chemical changes. We could study the atoms of the molecules, yet we would have to assume that the atoms are stable under nuclear fission. We could study the electrons and nucleons of the atoms, yet we would have to assume they are stable under weak nuclear forces.\footnote{This hierarchy of time scales is indeed used in many areas of science and engineering under the practice of multiscale modeling. \cite{Horstemeyer,Weinan} It allows one to break down a complex problem so that the relevant dynamics is studied efficiently at the right level of scale, gaining more insight and leading to faster calculations. Each scale will use different models and techniques, such as continuum mechanics, Monte-Carlo methods for statistical physics, molecular mechanics, quantum chemistry and so on. \cite{Brandt} During a computation cycle, each level of scale treats the faster and slower dynamics as fixed parameters, which are updated appropriately for the following iteration.} As we move to a finer and finer description, we have faster and faster scale processes that we assume do not change the basic constituents. At each level of the description what is assumed to be in equilibrium changes and what the laws describe changes as well. Yet, the overall structure remains the same: we have a timescale $dt$ such that within each interval $[t-dt, t]$ there is a process $\mathcal{p}_t$ such that the state $\mathcal{s}_t$ is an equilibrium of $\mathcal{p}_t$. The evolution $\mathcal{E}$, then, is quasistatic in terms of the processes in $\mathcal{P}$. That is, \textbf{all deterministic laws are quasistatic evolutions in terms of system-environment interaction processes}.

We have reached, then, the second major point of this article: \textbf{the mathematical structure we used to rearrange quantum mechanics is a general structure that applies to all systems} and as such we call it the \emph{fundamental model of physics}. It is fundamental because it always applies and what changes case by case are the specifics. In all cases we are going to have a set of states that are equilibria of faster scale processes. Each faster scale process will be, in general, non-deterministic in terms of the state and will act on the space of probability distributions. The law of evolution will be a quasistatic process in terms of the faster scale processes.

In classical mechanics, the state of a ball will be given by the position and momentum of the center of mass, which will be defined over a short time scale by integrating the jiggling created by the random collisions of air particles and internal motion of the ball's molecules. That averaging process will be the same regardless of the values of position and momentum; all states are equilibria of the same process and therefore $\mathcal{P}$ contains only one element: the identity map. Therefore, while the overall structure still exists, it is mathematically trivial and can be effectively disregarded as we concentrate on the law of evolution exclusively.

@@

We have seen that quantum mechanics fits within a more general framework that applies to all physical theories. States are always equilibria of faster scale processes, which in quantum mechanics are modeled with the projection. States are chosen to satisfy a deterministic law of evolution, necessarily describing a quasistatic process, which in quantum mechanics is modeled by the Schr\"{o}dinger equation. This general framework only uses standard physics concepts and puts no special role on the observer except the choice of what processes the system will be subjected to.

Violations of the Schr\"{o}dinger equation should be expected only for processes that cannot be considered quasistatic. If we were to implement a measurement process, then, in a way that the transfer of information between the system and the probe happens in a slow and controlled fashion, this could not provide a violation of the Schr\"{o}dinger equation, nor would it tell us that all projections are quasistatic processes. On the other hand, a better description of the faster scale processes given by the projections cannot be given in terms of unitary evolution, as this represents the quasistatic case. It would inevitably require a new state space as the standard quantum states can only describe equilibria of those processes. Moreover, a description at Planck time scale will ultimately fail to be continuous and quasistatic since a finite change in a finite time will not be divisible into infinitesimal adiabatic changes over infinitely many time steps.

To understand the suitable mathematical framework for those regimes where unitary evolution may not hold, we are proceeding by first identifying the physical requirements for all mathematical structures already used in physics, starting from the most basic ones like topological spaces (upon which is based all of differential geometry) and $\sigma$-algebras (upon which is based all of measure theory and probability theory). This long term effort is centered mainly around the development of an open access book\cite{Carc3} in which these issues can be explored in full detail and with a rigor more typical of mathematical disciplines. This rigor forces us to spell out the often hidden assumptions we have while constructing physical theories. By making explicit, for example, under which physically motivated circumstances time can be modeled by real numbers, or when states can be represented within Hilbert spaces, we believe we will be in a better position to understand what happens when those assumptions fail.

@@

\begin{thebibliography}{999}
	\bibitem[Carcassi(2018)]{Carc1}
	G. Carcassi, C. A. Aidala, D. J. Baker, L. Bieri; From physical assumptions to classical and quantum {Hamiltonian} and {Lagrangian} particle mechanics. {\em J. Phys. Comm.} {\bf 2018}, {\em 2(4)}, 045026.

	\bibitem[Genovese(2010)]{Genovese}
	M. Genovese; Interpretations of quantum mechanics and measurement problem. {\em Adv. Sci. Lett.} {\bf 2010}, {\em 3}, 249--258.

	\bibitem[Bassi(2013)]{Bassi}
	A. Bassi, K. Lochan, S. Satin, T. P. Singh, and H. Ulbricht; Models of wave-function collapse, underlying theories, and experimental tests. {\em Rev. Mod. Phys.} {\bf 2013}, {\em 85}, 471.

	\bibitem[Weinan(2011)]{Weinan}
	E. Weinan; Principles of Multiscale Modeling. {\em Princeton University} {\bf 2011}.

	\bibitem[Horstemeyer(2009)]{Horstemeyer}
	M. F. Horstemeyer; Multiscale Modeling: A Review. In Leszczyński, Jerzy; Shukla, Manoj K. (eds.). Practical Aspects of Computational Chemistry: Methods, Concepts and Applications. pp. 87–135. ISBN 978-90-481-2687-3.

	\bibitem[Brandt(2002)]{Brandt}
	Brandt A. (2002) Multiscale Scientific Computation: Review 2001. In: Barth T.J., Chan T., Haimes R. (eds) Multiscale and Multiresolution Methods. Lecture Notes in Computational Science and Engineering, vol 20. Springer, Berlin, Heidelberg

	\bibitem[Daneri(1966)]{Daneri}
	A. Daneri, A. Loinger and G. M. Prosperi; Further remarks on the relations between statistical mechanics and quantum theory of measurement. {\em Nuovo Cimento B} {\bf 1966}, {\em 44}, 119--128.

	\bibitem[Merlin(2015)]{Merlin}
	R. Merlin; A Heuristic Approach to the Quantum Measurement Problem: How to Distinguish Particle Detectors from Ordinary Objects. {\em Int. J. Mod. Phys. B} {\bf 2015}, {\em 29(22)}.

	\bibitem[Tamir(2013)]{Tamir}
	B, Tamir, E. Cohen; Introduction to Weak Measurements and Weak Values. {\em Quanta} {\bf 2013}, {\em 2}, 7–17.

	\bibitem[Carcassi(2019)]{Carc3}
	G. Carcassi, C. A. Aidala; Assumptions of Physics. {\em In preparation} {\bf 2019}, http://assumptionsofphysics/book.

\end{thebibliography}
